### 2021-11-08 manuscript.Rmd
final edits before making repo public

abstract: >
  Sequential analysis allows researchers to perform well-powered experiments with smaller sample sizes. In this practical guide, I outline the commonalities and differences between three sequential hypothesis testing procedures: group-sequential designs, the sequential Bayes Factor, and the Independent Segments Procedure. I illustrate how these sequential procedures can be implemented in a group-wise manner, and present two trade-offs of importance to the procedures: long-run error control vs. short-run evidence and efficiency in hypothesis testing vs. accuracy in parameter estimation. Using Monte Carlo simulations, I compare the performance of the procedures in terms of their error control (i.e., false-negative and false-positive decision rates), efficiency (i.e., reduction in required sample size), and accuracy (i.e., bias and variance in effect size estimation). I find that effect size estimates stemming from the Independent Segments Procedure are highly inaccurate, and systematically exclude the true population effect size. I conclude with practical recommendations for researchers who wish to utilize the benefits of sequential hypothesis testing while retaining their ability to obtain a credible effect size estimate.

keywords          : "sequential analysis, hypothesis testing, efficiency, estimation, bias"
wordcount         : "-"

@@

Because the OBF-like procedure sets such a strict $\alpha$ threshold for the first look, a researcher employing the O'Brien-Fleming-like procedure will have relatively low power, and consequently a low probability of stopping for efficacy, in the first look. Thus, when a researcher uses the O'Brien Fleming-like procedure and wishes to be well-powered during the earlier looks, it is often recommended to have the first look take place later in the data collection process (rather than spacing the looks equally; Daniël Lakens, personal communication, December 19, 2020). For example, when $k_{max} = 3$, a researcher can decide to equally space their looks when employing a Pocock-like design (i.e., analyze the data after 33%,  67%, and 100% of the data are collected, respectively), while taking unequally spaced looks when employing an O'Brien-Fleming-like design (e.g., analyze the data after 50%, 75%, and 100% of the data are collected, respectively).

Because the use of group-sequential procedures allows for early stopping, it generally reduces the average sample size needed (with certain exceptions; e.g., when the true effect size is equal to or close to zero, the use of sequential procedures tends to increase the average sample size needed). For example, if a researcher wants to conduct a one-sided, two-sample *t* test ($H_0:~\delta$ = 0 against $H_1:~\delta > 0$) with 80% power to detect a Cohen's *d* (i.e., a standardized mean difference) of 0.5, a conventional Null Hypothesis Significance Test with an overall $\alpha$ of 5% would require `r n_fixed` participants per group. Using the Pocock-like GS design outlined above, the same study can be conducted with only `r n_p` participants per group per look. Using the O'Brien-Fleming like design outlined above, this same study can be conducted with `r n_of[1]` participants per group during the first look, and `r n_of[2]` participants per group during the last two looks. Below, I briefly outline the decision-theoretic procedure of GS designs.

@@

## *Independent segments procedure* 
Another sequential procedure recently proposed for use in psychological research is the Independent Segments Procedure [ISP, @miller2020]. A central property of the ISP is that data is collected in a series of independent experiments ('segments'). Using other sequential procedures (such as the GS design and the Sequential Bayes Factor), data collection and analysis is cumulative: when an experiment is continued (because the interim data did not cross the pre-specified boundaries for early stopping), a researcher collects more observations, adds these to the observations from previous looks, and calculates an interim test statistic on the basis of all the data collected so far. In contrast, when an independent segments experiment is continued, the researcher discards the previous batch of observations and calculates an interim test statistic on the basis of a new, completely independent, batch of observations. Unlike the SBF and GS-designs, the ISP ignores data from earlier looks. The ISP, by design, incorporates information loss.

@@

Second, the ISP only applies to research contexts in which a single, key hypothesis is tested [@miller2020]. This, unfortunately, is unusually restrictive to the settings in which sequential hypothesis testing is common. For example, medical trials often have multiple "*endpoints*" that need to be monitored at the same time: a primary endpoint (e.g., treatment success or survival) and multiple secondary -- but exceedingly important -- efficacy or safety endpoints [@wassmer2016]. An enlightening example can be found in the FDA approval process for carvedilol (a drug developed for congestive heart failure), in which the endpoints were -- among other things -- mortality, distance that can be walked on a treadmill, exercise tolerance, cardiovascular hospitalizations, quality of life, and the patient and physician's global assessment [@fisher1999a].

@@

Of note, because the properties of group-sequential designs are well-studied and understood, bias correction procedures are now widely available for group-sequential designs. That is, rather than relying on a naive, uncorrected effect size estimate stemming from a sequential design (e.g., Cohen's *d*), researchers can use a median unbiased estimator (which can be easily implemented in widely used sequential analysis software such as `rpact`). A similar bias-correction procedure can be followed for the Sequential Bayes Factor: instead of relying on a naive, uncorrected effect size estimate, a researcher employing a SBF design can use the mean of the posterior distribution as the effect size estimator, which shrinks the estimate in early terminations and thus counteracts the effect size inflation caused by early stopping [@schonbrodt2017]. No such bias correction procedure, however, exists for the Independent Segments Procedure. Thus, it is of great interest to study the accuracy of effect size estimation when relying on the Independent Segments Procedure.

@@

@schonbrodt2017 show that a highly flexible implementation of the Sequential Bayes Factor -- i.e., the SBF in all its glory, with as many looks as is desired -- is at least as efficient as a default group-sequential design (i.e., a group-sequential approach that does not incorporate binding futility bounds on the basis of a $\beta$-spending approach). In this paper, I implement the SBF with $k_{max}$ equally spaced looks, for two reasons: (a) to allow for maximum comparability between the sequential procedures under discussion, controlling for as many confounds as possible; and (b) to account for practical considerations: it may very well be impractical for researchers to implement the maximum-flexibility SBF approach and analyze data after every participant. Thus, I focus on a slightly different, pragmatic question: For a fixed number of looks at data (i.e., a "group-wise" implementation), how do these respective procedures perform in terms of their error rates, efficiency, and accuracy?

@@

For maximum comparability to the sequential procedures that allow some principled form of support for the null hypothesis (i.e., the Sequential Bayes Factor and the Group-Sequential Designs with binding futility bounds), we combine the fixed-sample hypothesis test with an equivalence test. Conventionally, the Null Hypothesis Significance test is not able to distinguish between failures to reject the null hypothesis due to evidence of absence or absence of evidence. Pairing the fixed sample hypothesis test with an equivalence test overcomes this limitation, by allowing us to reject the presence of a difference more extreme than a smallest effect size of interest [@lakens2018]. In the following section, I briefly outline how each property under study -- error rates, efficiency, and accuracy in effect size estimation -- is evaluated.

@@

Procedural efficiency is evaluated by means of the average (expected) sample size across several true effect sizes for each procedure. Accuracy in effect size estimation is evaluated by means of the density of empirical effect size estimates, the bias in effect size estimates (i.e., the median estimated effect size compared to the true effect size), and the mean squared error of effect size estimates (i.e., the sum of the squared bias and the variance of effect size estimates) derived from each procedure. For all procedures, accuracy is evaluated for the obtained, uncorrected effect size estimates --- in addition, for procedures that have a bias-adjusted estimator available (i.e., the Sequential Bayes Factor and the group-sequential designs), I also evaluate the accuracy of the bias-corrected estimates.

@@

I simulate populations with a specific standardized mean difference $\delta$ and examine the efficiency (i.e., average sample size required), error rates (i.e., rates of false-positive and false-negative evidence), and accuracy (i.e., of the simulation estimate of the population effect size) of five hypothesis testing procedures: the fixed-sample Neyman-Pearson hypothesis test paired with an equivalence test [@lakens2014], the Independent Segments Procedure [@miller2020], a group-sequential design with Pocock-like $\alpha$- and $\beta$-spending functions [@pocock1977; @wassmer2016], a group-sequential design with O'Brien-Fleming-like $\alpha$- and $\beta$-spending functions [@fleming1984; @wassmer2016], and the sequential Bayes Factor [@schonbrodt2017].

To facilitate discussion, I focus on one typical scenario in reporting the simulation results. Full results that vary additional parameters and reproducible analysis scripts are available at [link]. For the frequentist procedures (i.e., the fixed-sample N-P procedure paired with an equivalence test, the ISP, and the Pocock-like and O'Brien-Fleming-like GS designs, respectively) under discussion, the chosen typical scenario corresponds to a one-tailed, two-sample hypothesis test powered to 80% ($\beta$ = 0.2) to detect a hypothesized population effect size of $\delta$ = 0.5 with $\alpha$ = .05. In the fixed-sample case, these specifications require a sample size of `r n_fixed` observations per group. The equivalence bounds to achieve 80% power with N = `r n_fixed` are a lower equivalence bound of $\Delta_L$ = -0.58 and an upper equivalence bound of $\Delta_U$ = 0.58.

All sequential procedures, with the exception of the O'Brien-Fleming-like GS design, are set to a maximum number of 3 equally spaced looks (i.e., the interim data are looked at twice, after 33% and 67% of the data are collected, and the final look takes place when all data are collected). In contrast, because the O'Brien-Fleming-like GS design sets such a stringent $\alpha$ threshold in the first look, it is often recommended to have the first look take place on a larger fraction of the data (Daniël Lakens, personal communication, December 19, 2020). In this case, we implemented the O'Brien-Fleming-like GS design with the two interim looks taking place after 50% and 75% of the data, respectively, are collected.

@@

## *Fixed-sample hypothesis test* 
  *
1. Randomly sample $N_{text{fixed}}$ observations from each of the two specified populations
2. Run a one-tailed, two-sample *t* test on the mean difference between the two samples
  a. If $p \le \alpha$: decide to reject $H_0$
  b. If $p > \alpha$: decide not to reject $H_0$
3. Run a one-tailed, two-sample equivalence test (TOST) on the mean difference between the two samples
  a. If $p_{\text{TOST},1:2} \le \alpha$: decide to reject $H_1$ 
  b. If $p_{\text{TOST},1:2} > \alpha$: decide not to reject $H_1$ (inconclusive)
4. To estimate the standardized effect size, calculate Cohen's *d* (the difference in sample means divided by their pooled standard deviation)
5. Repeat 10,000 times

@@

All data, analysis code, and research materials are available at [link]. The study was not preregistered. This manuscript was created using `r cite_r("r-refs.bib")`.

@@

The error rates involved in the procedures under discussion are shown in Table 2A, Table 2B, and Figure 4. Table 2A describes the rates of True Negative, Inconclusive, and False Positive results when the true population effect size $\delta$ = 0; Table 2B describes the rates of True Positive, Inconclusive, and False Negative results when the true population effect size $\delta$ = 0.50. As can be read from the tables, the frequentist procedures all perform as expected: 5% of simulations rejected the null hypothesis when $\delta$ = 0 (i.e., a type I error rate of 5%; see the False Positive column Table 2A) and 80-82% of simulations rejected the null hypothesis when $\delta$ = 0.5 (i.e., a type II error rate of 18-20%; see the True Positive column of Table 2B).

@@

Table 3 and Figure 5 show the average sample size required for the procedures under investigation, as a function of the true population effect size. As discussed, the fixed-sample procedure requires 51 participants per group. As can be read from Table 3 and Figure 5, the sequential procedures allow for a reduction in the average sample size required. Across the procedures, efficiency gains appear to be rather similar. When the true population effect size is relatively small, the max-SBF procedure is less efficient than its frequentist counterparts. When the true population effect size is relatively large, the O'Brien-Fleming approach is slightly less efficient than its counterparts. This can be explained by the relatively stringent $\alpha_1$ levels of the O'Brien-Fleming procedure in earlier segments (recall that, for the O'Brien-Fleming-like procedure, $\alpha_{1,1}$ = `r alpha_of[1]` and $\alpha_{1,2}$ = `r alpha_of[2]`). Across all true effect sizes, the Pocock-like procedure, with our chosen parameters, provides the greatest efficiency gains.

@@

In figures A1-A4 in Appendix A, the reader can find that this observation is not merely idiosyncratic to the test chosen (a two-sample *t* test with $\alpha$ = .05 and $\beta = .20$ for a hypothesized population effect size $\delta$ = 0.5). In contrast, across dozens of different test specifications, effect size estimates stemming from the ISP are consistently bimodal and systematically exclude the true population effect size from being observed. Because the ISP involves running three independent mini-experiments (rather than accumulating data over time), the individual segments -- with the exception of the last segment, for which a very weak $\alpha$ is set -- are severely underpowered. The first $k_{max} - 1$ segments of the ISP tend to have less than 50% power. As a result, the critical effect size for these segments will always be higher than the hypothesized effect size. In our worked example, the first two individual segments have a power of `r round(isp_power*100)`%. To make sure that the hypothesized effect size is not consistently weeded out by the procedure's significance filter, the power of each individual segment must be set to at least 50%.

It is enlightening to consider that the Sequential Bayes Factor and Group-Sequential procedures are similarly biased in earlier segments. Early stopping acts as a filter that weeds out moderately sized effects: the only effect sizes that are maintained are the ones that are relatively small or in the wrong direction (e.g., when stopping for futility), or the ones that are really large (e.g., when stopping for efficacy). However, due to the cumulative nature of data collection and analysis, the bias inherent in early terminations tends to be compensated for by studies that terminate at later stages [@goodman2007; @fan2004; @schonbrodt2017]. As more observations are collected, the variability in effect size estimates decreases. Unfortunately, this is not the case for the Independent Segments Procedure.

The median effect size estimates across all possible values of $\delta$ are shown in Table 4. As expected, the median effect size estimates stemming from fixed-sample hypothesis tests are equal to the true population effect size. The uncorrected median effect size estimates stemming from the Pocock-like and O'Brien-Fleming-like designs deviate only slightly from the true population effect size. When the population effect size is relatively small ($\delta < 0.2$), the Pocock-like and O'Brien-Fleming-like effect size estimates slightly underestimate the true effect; when the population effect size is relatively large ($\delta > 0.4$), the Pocock-like and O'Brien-Fleming-like effect size estimates slightly overestimate the true population effect size. As can be read in the Table, the group-sequential median-unbiased estimator, used to adjust for bias in the effect size estimates, provides a remarkable improvement in the accuracy of the parameter estimates. In all cases, the group-sequential bias-adjusted effect size estimate is equal to or no more than $d = 0.02$ removed from the true population parameter.

For the unadjusted Bayesian effect size estimates, a similar patterns holds: smaller effect sizes are slightly underestimated, while larger effect sizes are slightly overestimated. The bias-adjusted estimate (the mean of the posterior distribution) increases the estimates for smaller population effect sizes (i.e., for early terminations in which the null was accepted), and shrinks the estimates for larger population effect sizes (i.e., for early terminations in which the alternative was accepted). Although this provides a slight improvement in the parameter estimate for certain population effect sizes ($\delta$ = 0.2, 0.5, and 0.6), the Bayesian bias-corrected estimates do not perform as well as the group-sequential bias-corrected estimates.

@@

Figure 7 visualizes the information captured by Table 4, illustrating the differences between the median estimated effect sizes and the true effect sizes. For the Sequential Bayes Factor, we can see that the uncorrected estimated slightly underestimate the true population effect size when it is small, and slightly overestimate the true population effect size when it is large. The bias-adjusted estimate, the mean of the posterior distribution, corrects for this phenomenon -- but it is important to note that the Bayesian bias-adjusted estimates only provide a slight improvement or, for certain parameters, an overcorrection. Contrast this with the two bottom row panels for the group-sequential designs. Here, we witness that the group-sequential median unbiased estimator lives up to its name. As before, the ISP shows the greatest overall deviation from the true effect size.

@@

Figure 8 shows the mean squared error of the five procedures for the naive effect size estimates and the available bias-adjusted estimates. The mean squared error is a measure of the accuracy of an estimator, composed of two elements: variance (the spread of effect size estimates from one sample to the other) and squared bias (the difference between the mean estimated effect size and the true population effect size). We find that the Independent Segments Procedure has the greatest degree of mean squared error. In addition, we find that the error stems mostly from the higher variance of effect size estimates. As discussed before, the variance in the effect size estimates stemming from the ISP is caused by the process of conducting three independent mini-experiments, rather than a cumulative study. In the language of decision theory, the ISP is inadmissible: its total squared error risk exceeds that of other hypothesis testing procedures (in our worked example of $\delta$ = 0.5, MSE = `r mean_mse[1]` for the ISP, compared to `r mean_mse[2]` for the Pocock-like design, `r mean_mse[3]` for the O'Brien-Fleming design, `r mean_mse[4]` for the SBF, and `r mean_mse[5]` for the fixed-sample procedure).

To provide more meat to this argument, we separate the error of the procedures by segment in Figure 9. Here, we can clearly see that stopping after a first look causes error for all sequential procedures. For both the SBF and the Pocock-like procedure, however, error becomes near-negligible in the second and third segment. As a result, overall error (across all segments) is much lower. This is akin to the bias compensation process we have described above. For the Independent Segments Procedure, as we now know, error remains substantial in every segment, leading to non-negligible overall error.

@@

In this final section, I present several practical considerations for researchers who wish to choose between the sequential procedures under discussion. Group-sequential designs, Sequential Bayes Factors, and the Independent Segments Procedure each have advantages and disadvantages. In deciding between sequential hypothesis testing procedures, researchers need to carefully weigh the advantages and disadvantages of the procedures of interest, with regard to the context in which they are conducting their research [@stefan2020].

>>>>

abstract: >
  Sequential analysis allows researchers to perform well-powered experiments with smaller sample sizes. I outline the commonalities and differences between three sequential hypothesis testing procedures: group-sequential designs, the sequential Bayes Factor, and the Independent Segments Procedure. I illustrate how these sequential procedures can be implemented in a group-wise manner, and present two trade-offs of importance to the procedures: long-run error control vs. short-run evidence and efficiency in hypothesis testing vs. accuracy in parameter estimation. Using Monte Carlo simulations, I compare the performance of the procedures in terms of their error control (i.e., false-negative and false-positive decision rates), efficiency (i.e., reduction in required sample size), and accuracy (i.e., bias and variance in effect size estimation). I find that effect size estimates stemming from the Independent Segments Procedure are highly inaccurate, and systematically exclude the true population effect size. I conclude with practical recommendations for researchers who wish to utilize the benefits of sequential hypothesis testing while retaining their ability to obtain a credible effect size estimate.

keywords          : "sequential analysis, hypothesis testing, efficiency, estimation, bias"
wordcount         : "11,387"

@@

Because the OBF-like procedure sets such a strict $\alpha$ threshold for the first look, a researcher employing the O'Brien-Fleming-like procedure will have relatively low power, and consequently a low probability of stopping for efficacy, during the first look. Thus, when a researcher uses the O'Brien Fleming-like procedure and wishes to be well-powered during the earlier looks, it is often recommended to have the first look take place later in the data collection process (rather than spacing the looks equally; Daniël Lakens, personal communication, December 19, 2020). For example, when $k_{max} = 3$, a researcher can decide to space their looks equally when employing a Pocock-like design (i.e., analyze the data after 33%,  67%, and 100% of the data are collected, respectively), while taking unequally spaced looks when employing an O'Brien-Fleming-like design (e.g., analyze the data after 50%, 75%, and 100% of the data are collected, respectively).

Because the use of group-sequential procedures allows for early stopping, it generally reduces the average sample size needed (with certain exceptions; e.g., when the true effect size is equal to or close to zero, the use of sequential procedures can increase the average sample size needed). For example, if a researcher wants to conduct a one-sided, two-sample *t* test ($H_0:~\delta$ = 0 against $H_1:~\delta > 0$) with 80% power to detect a Cohen's *d* (i.e., a standardized mean difference) of 0.5, a conventional Null Hypothesis Significance Test with an overall $\alpha$ of 5% would require `r n_fixed` participants per group. Using the Pocock-like GS design outlined above, the same study can be conducted with only `r n_p` participants per group per look. Using the O'Brien-Fleming like design outlined above, this same study can be conducted with `r n_of[1]` participants per group during the first look, and `r n_of[2]` participants per group during the last two looks. Below, I briefly outline the decision-theoretic procedure of GS designs.

@@

## *Independent segments procedure* 
Another sequential procedure recently proposed for use in psychological research is the Independent Segments Procedure [ISP\; @miller2020]. A central property of the ISP is that data is collected in a series of independent experiments ('segments'). Using other sequential procedures (such as the GS designs and the Sequential Bayes Factor), data collection and analysis is cumulative: when an experiment is continued (because the interim data did not cross the pre-specified boundaries for early stopping), a researcher collects more observations, adds these to the observations from previous looks, and calculates an interim test statistic on the basis of all the data collected so far. In contrast, when an independent segments experiment is continued, the researcher discards the previous batch of observations and calculates an interim test statistic on the basis of a new, completely independent, batch of observations. Unlike the SBF and GS-designs, the ISP ignores data from earlier looks. The ISP, by design, incorporates information loss.

@@

Second, the ISP only applies to research contexts in which a single, key hypothesis is tested [@miller2020]. This, unfortunately, is unusually restrictive to the settings in which sequential hypothesis testing is common. For example, medical trials often have multiple "*endpoints*" that need to be monitored at the same time: a primary endpoint (e.g., treatment success or survival) and multiple secondary -- but exceedingly important -- efficacy or safety endpoints [@wassmer2016].

@@

Of note, because the properties of group-sequential designs are well-studied and understood, bias correction procedures are now widely available for group-sequential designs. That is, rather than relying on a naive, uncorrected effect size estimate stemming from a sequential design (e.g., Cohen's *d*), researchers can use a median unbiased estimator [which can be easily implemented in widely used sequential analysis software such as `rpact`\; @R-rpact]. A similar bias-correction procedure can be followed for the Sequential Bayes Factor: instead of relying on a naive, uncorrected effect size estimate, a researcher employing a SBF design can use the mean of the posterior distribution as the effect size estimator, which shrinks the estimate in early terminations and thus counteracts the effect size inflation caused by early stopping [@schonbrodt2017]. No such bias correction procedure, however, exists for the Independent Segments Procedure. Thus, it is of great interest to study the accuracy of effect size estimation when relying on the Independent Segments Procedure.

@@

@schonbrodt2017 show that a highly flexible implementation of the Sequential Bayes Factor -- i.e., the SBF in all its glory, with as many looks as is desired -- is at least as efficient as a default group-sequential design (i.e., a group-sequential approach that does not incorporate binding futility bounds on the basis of a $\beta$-spending approach). In this paper, I offer a more balanced comparison between the two procedures. First, because the group-sequential approach also offers a principled decision rule to reject an alternative hypothesis, I implement the group-sequential designs with binding futility bounds. Second, I implement the SBF with $k_{max}$ equally spaced looks, for two reasons: (a) to allow for maximum comparability between the sequential procedures under discussion, and (b) to account for practical considerations; it may very well be impractical for researchers to implement the maximum-flexibility SBF approach and analyze data after every participant. Thus, I focus on a slightly different, pragmatic question: For a fixed number of looks at data (i.e., a "group-wise" implementation), how do these respective procedures perform in terms of their error rates, efficiency, and accuracy?

@@

For maximum comparability to the sequential procedures that allow some principled form of support for the null hypothesis (i.e., the Sequential Bayes Factor and the group-sequential designs with binding futility bounds), we combine the fixed-sample hypothesis test with an equivalence test. Conventionally, the Null Hypothesis Significance test is not able to distinguish between failures to reject the null hypothesis due to evidence of absence or absence of evidence. Pairing the fixed sample hypothesis test with an equivalence test overcomes this limitation, by allowing us to reject the presence of a difference more extreme than a smallest effect size of interest [@lakens2018]. In the following section, I briefly outline how each property under study -- error rates, efficiency, and accuracy in effect size estimation -- is evaluated.

@@

Procedural efficiency is evaluated by means of the average (expected) sample size across several true effect sizes for each procedure. Accuracy in effect size estimation is evaluated by means of the density of empirical effect size estimates, the bias in effect size estimates (i.e., the median estimated effect size compared to the true effect size), and the mean squared error of effect size estimates (i.e., the sum of the squared bias and the variance of effect size estimates) derived from each procedure. For all procedures, accuracy is evaluated for the obtained, uncorrected effect size estimates. In addition, for procedures that have a bias-adjusted estimator available (i.e., the Sequential Bayes Factor and the group-sequential designs), I also evaluate the accuracy of the bias-corrected estimates.

@@

I simulate populations with a specific standardized mean difference $\delta$ and examine the efficiency (i.e., average sample size required), error rates (i.e., rates of false-positive and false-negative evidence), and accuracy (i.e., of the simulation estimate of the population effect size) of five hypothesis testing procedures: the fixed-sample Neyman-Pearson hypothesis test paired with an equivalence test [@lakens2014], a group-sequential design with Pocock-like $\alpha$- and $\beta$-spending functions [@pocock1977; @wassmer2016], a group-sequential design with O'Brien-Fleming-like $\alpha$- and $\beta$-spending functions [@fleming1984; @wassmer2016], the sequential Bayes Factor [@schonbrodt2017], and the Independent Segments Procedure [@miller2020].

To facilitate discussion, I focus on one typical scenario in reporting the simulation results. For the frequentist procedures (i.e., the fixed-sample N-P procedure paired with an equivalence test, the ISP, and the Pocock-like and O'Brien-Fleming-like GS designs, respectively) under discussion, the chosen typical scenario corresponds to a one-tailed, two-sample hypothesis test powered to 80% ($\beta$ = 0.2) to detect a hypothesized population effect size of $\delta$ = 0.5 with $\alpha$ = .05. In the fixed-sample case, these specifications require a sample size of `r n_fixed` observations per group. The equivalence bounds to achieve 80% power with N = `r n_fixed` are a lower equivalence bound of $\Delta_L$ = -0.58 and an upper equivalence bound of $\Delta_U$ = 0.58.

All sequential procedures, with the exception of the O'Brien-Fleming-like GS design, are set to a maximum number of 3 equally spaced looks (i.e., the interim data are looked at twice, after 33% and 67% of the data are collected, and the final look takes place when all data are collected). In contrast, because the O'Brien-Fleming-like GS design sets such a stringent $\alpha$ threshold in the first look, it is often recommended to have the first look take place on a larger fraction of the data (Daniël Lakens, personal communication, December 19, 2020). In this case, I implemented the O'Brien-Fleming-like GS design with the two interim looks taking place after 50% and 75% of the data, respectively, are collected.

@@

## *Fixed-sample hypothesis test* 
  *
1. Randomly sample $N_{\text{fixed}}$ observations from each of the two specified populations
2. Run a one-tailed, two-sample *t* test on the mean difference between the two samples
  a. If $p \le \alpha$: decide to reject $H_0$
  b. If $p > \alpha$: decide not to reject $H_0$
3. Run a one-tailed, two-sample equivalence test (TOST) on the mean difference between the two samples
  a. If $p_{\text{TOST},1:2} \le \alpha$: decide to reject $H_1$ 
  b. If $p_{\text{TOST},1:2} > \alpha$: decide not to reject $H_1$ (inconclusive)
4. To estimate the standardized effect size, calculate Cohen's *d* (the difference in sample means divided by their pooled standard deviation)
5. Repeat 10,000 times

@@

All data, code, and research materials are available at [link]. The study was not preregistered. This manuscript was created using `r cite_r("r-refs.bib")`.

@@

The error rates involved in the procedures under discussion are shown in Table 2A, Table 2B, and Figure 4. Table 2A describes the rates of True Negative, Inconclusive, and False Positive results when the true population effect size $\delta$ = 0; Table 2B describes the rates of True Positive, Inconclusive, and False Negative results when the true population effect size $\delta$ = 0.50. Figure 4 shows the True Positive, Inconclusive, and False Negative rates for all population effect sizes greater than zero ($\delta$ = 0.2, 0.4, 0.6, 0.8, 1). As can be read from the tables, the frequentist procedures all perform as expected: 5% of simulations rejected the null hypothesis when $\delta$ = 0 (i.e., a type I error rate of 5%; see the False Positive column Table 2A) and 80-82% of simulations rejected the null hypothesis when $\delta$ = 0.5 (i.e., a type II error rate of 18-20%; see the True Positive column of Table 2B).

@@

Table 3 and Figure 5 show the average sample size required for the procedures under investigation, as a function of the true population effect size. As discussed, the fixed-sample procedure requires 51 participants per group. As can be read from Table 3 and Figure 5, the sequential procedures allow for a reduction in the average sample size required. Across the procedures, efficiency gains appear to be rather similar. When the true population effect size is relatively small, the max-SBF procedure is less efficient than its frequentist counterparts. When the true population effect size is relatively large, the O'Brien-Fleming approach is slightly less efficient than its counterparts. This can be explained by the relatively stringent $\alpha_1$ levels of the O'Brien-Fleming procedure in the first segment (recall that, for the O'Brien-Fleming-like procedure, $\alpha_{1,1}$ = `r alpha_of[1]`). Across all true effect sizes, the Pocock-like procedure, with our chosen parameters, provides the greatest efficiency gains.

@@

In figure A1 in Appendix A, the reader can find that this observation is not merely idiosyncratic to the test chosen (a two-sample *t* test with $\alpha$ = .05 and $\beta = .20$ for a hypothesized population effect size $\delta$ = 0.5). In contrast, across dozens of different test specifications, effect size estimates stemming from the ISP are consistently bimodal and systematically exclude the true population effect size from being observed. Because the ISP involves running three independent mini-experiments (rather than accumulating data over time), the individual segments -- with the exception of the last segment, for which a very weak $\alpha$ is set -- are severely underpowered. The first $k_{max} - 1$ segments of the ISP tend to have less than 50% power. As a result, the critical effect size for these segments will always be higher than the hypothesized effect size. In our worked example, the first two individual segments have a power of `r round(isp_power*100)`%. To make sure that the hypothesized effect size is not consistently weeded out by the procedure's significance filter, the power of each individual segment must be set to at least 50%.

It is enlightening to consider that the Sequential Bayes Factor and Group-Sequential procedures are similarly biased in earlier segments. Early stopping acts as a filter that weeds out moderately sized effects: the only effect sizes that are maintained are the ones that are relatively small or in the wrong direction (e.g., when stopping for futility), or the ones that are really large (e.g., when stopping for efficacy). However, due to the cumulative nature of data collection and analysis, the bias inherent in early terminations tends to be compensated for by studies that terminate at later stages [@goodman2007; @fan2004; @schonbrodt2017]. As more observations are collected, the variability in effect size estimates decreases. In meta-analyses of findings stemming from cumulative sequential procedures (such as the SBF and the GS designs), less weight will be given to the relatively biased findings stemming from earlier terminations and -- because of their larger sample sizes -- more weight will be given to the relatively unbiased findings stemming from later terminations. As a result, the meta-analytic bias in the effect size estimates becomes negligible. Unfortunately, this is not the case for the Independent Segments Procedure. For the Independent Segments Procedure,  the first $k_{max} - 1$ segments are highly biased due to their low power, which only allows very small or very large effects to filter through. The last segment has slightly higher power because a weaker $\alpha$ is set (although this increase in power is less substantial than for cumulative procedures, which receive a marked increase in power due to their accumulating sample size as the study progresses), and thus tends to be slightly less biased. However, because each independent segment has equal sample sizes, they are all given equal weight. As a result, the overall bias in the effect size estimates remains severe.

The median effect size estimates across all possible values of $\delta$ are shown in Table 4. As expected, the median effect size estimates stemming from fixed-sample hypothesis tests are equal to the true population effect size. The uncorrected median effect size estimates stemming from the Pocock-like and O'Brien-Fleming-like designs deviate only slightly from the true population effect size. When the population effect size is relatively small ($\delta < 0.2$), the Pocock-like and O'Brien-Fleming-like effect size estimates slightly underestimate the true effect; when the population effect size is relatively large ($\delta > 0.4$), the Pocock-like and O'Brien-Fleming-like effect size estimates slightly overestimate the true population effect size. As can be read in the Table, the group-sequential median-unbiased estimator, used to adjust for bias in the effect size estimates, provides a remarkable improvement in the accuracy of the parameter estimates. In all cases, the group-sequential bias-adjusted effect size estimate is equal to or no more than $d$ = 0.02 removed from the true population parameter.

For the unadjusted Bayesian effect size estimates, a similar patterns holds: smaller effect sizes are slightly underestimated, while larger effect sizes are slightly overestimated. The bias-adjusted estimate (the mean of the posterior distribution) increases the estimates for smaller population effect sizes (i.e., for early terminations in which the null was accepted), and shrinks the estimates for larger population effect sizes (i.e., for early terminations in which the alternative was accepted). Although this provides a slight improvement in the parameter estimate for certain population effect sizes ($\delta$ = 0.2, 0.5, and 0.6), the Bayesian bias-corrected estimates do not perform as well as the group-sequential bias-corrected estimates. The mean of the posterior distribution either slightly improves or, for certain parameters, overcorrects the unadjusted Bayesian effect size estimates.

@@

In our simulations, we find that -- for moderate population parameters (0.4 < $\delta$ < 0.6) -- the ISP (powered to detect an effect of $\delta$ = 0.5, with $1 - \beta = 0.8$) severely overestimates effect sizes. In Figure A4 in Appendix A, the reader can find that this observation generalizes across many different specifications of statistical power and hypothesized effect size. Due to the low power of each independent segment, the critical effect size is higher -- thus, observed significant effects are likely to exaggerate the true population parameter. In Figure A4, we find that this exaggeration is severe. For example, when $\delta = 0.6$ and power is 80%, the probability of observing a significant effect that is more than 1.5 times the size of the true population parameter (i.e., the probability of observing *d* > 0.9) is 45%, compared to only 15% for a fixed sample hypothesis test.

Figure 7 visualizes the information captured by Table 4, illustrating the differences between the median estimated effect sizes and the true effect sizes. For the Sequential Bayes Factor, we can see that the uncorrected estimated slightly underestimate the true population effect size when it is small, and slightly overestimate the true population effect size when it is large. The bias-adjusted estimate, the mean of the posterior distribution, corrects for this phenomenon -- but, as noted before, the Bayesian bias-adjusted estimates only provide a slight improvement or, for certain parameters, an overcorrection. Contrast this with the two bottom row panels for the group-sequential designs. Here, we witness that the group-sequential median unbiased estimator lives up to its name. As before, the ISP shows the greatest overall deviation from the true effect size.

@@

Figure 8 shows the mean squared error of the five procedures for the naive effect size estimates and the available bias-adjusted estimates. The mean squared error is a measure of the accuracy of an estimator, composed of two elements: variance (the spread of effect size estimates from one sample to the other) and squared bias (the difference between the mean estimated effect size and the true population effect size). We find that the Independent Segments Procedure has the greatest degree of mean squared error. In addition, we find that the error stems mostly from the higher variance of effect size estimates. As discussed before, the variance in the effect size estimates stemming from the ISP is caused by the process of conducting three independent mini-experiments, rather than a cumulative study.

In Figures A2 and A3 in Appendix A, the reader can observe that our findings regarding the bias, variance, and mean squared error of the Independent Segments Procedure generalize across many different test specifications. In the language of decision theory, the ISP is inadmissible: its total squared error risk exceeds that of other hypothesis testing procedures (in our worked example of $\delta$ = 0.5, MSE = `r mean_mse[1]` for the ISP, compared to `r mean_mse[2]` for the Pocock-like design, `r mean_mse[3]` for the O'Brien-Fleming design, `r mean_mse[4]` for the SBF, and `r mean_mse[5]` for the fixed-sample procedure).

To provide more meat to the argument that the error of the ISP is caused by the process of conducting three independent mini-experiments, we separate the error of the procedures by segment in Figure 9. Here, we can clearly see that stopping after a first look causes error for all sequential procedures. For both the SBF and the group-sequential procedures, however, error becomes near-negligible in the second and third segment. As a result, overall error (across all segments) is much lower. This is akin to the bias compensation process we have described above. For the Independent Segments Procedure, as we now know, error remains substantial in every segment, leading to non-negligible overall error. We can see that the error for the first $k_{max} -1$ segments is highest, while the error for the last segment is slightly decreased. As discussed before, this is due to the slightly higher power of the last segment, for which a weaker $\alpha$ is set. This error reduction, however, simply does not compare to the reduction that would have been obtained had the experiment been cumulative.

@@

In this final section, I present several practical considerations for researchers who wish to choose between the sequential procedures under discussion. Group-sequential designs, Sequential Bayes Factors, and the Independent Segments Procedure each have advantages and disadvantages. In deciding between sequential hypothesis testing procedures, researchers need to carefully weigh the advantages and disadvantages of the procedures of interest, with regard to the context in which they are conducting their research [@stefan2020]. Table 1 presents several of the commonalities and differences between the sequential hypothesis testing procedures.
